---
title: "Exploratory Data Analysis (EDA)"
author: "Orły"
date: "`r Sys.Date()`"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE)
```

# Introduction

This notebook conducts an **Exploratory Data Analysis (EDA)** on the `richmondway.csv` dataset, which contains episode-level statistics related to the character Roy Kent in a TV series.

------------------------------------------------------------------------

## 1. Data Loading and Inspection

```{r}
library(tidyverse)
library(skimr)
library(visdat)
library(tidymodels)

set.seed(44)

df <- read_csv("richmondway.csv")

dim(df)
glimpse(df)
skim(df)
```

------------------------------------------------------------------------

## 2. Missing Values Overview

```{r fig.width=8, fig.height=5}
vis_miss(df, cluster = TRUE) +
  labs(title = "Missing Values")
```

------------------------------------------------------------------------

## 3. Data Cleaning and Preparation

```{r}
df <- df %>%
  mutate(
    Season = as.integer(Season),
    Episode = as.integer(Episode),
    F_score = as.numeric(F_score),
    F_perc = as.numeric(F_perc),
    year = 2020 + Season
  )

sum(duplicated(df))
```

------------------------------------------------------------------------

## 4. Univariate Analysis

### Distribution of F_score

```{r fig.width=10}
p1 <- ggplot(df, aes(x = F_score)) +
  geom_histogram(bins = 15, fill = "skyblue", color = "black") +
  labs(title = "Histogram of F_score", x = "F_score", y = "Frequency")

p2 <- ggplot(df, aes(y = F_score)) +
  geom_boxplot(fill = "skyblue") +
  labs(title = "Boxplot of F_score", y = "F_score")

p1
p2
```

### Distribution of F_perc

```{r fig.width=10}
p1 <- ggplot(df, aes(x = F_perc)) +
  geom_histogram(bins = 15, fill = "lightcoral", color = "black") +
  labs(title = "Histogram of F_perc", x = "F_perc (%)", y = "Frequency")

p2 <- ggplot(df, aes(y = F_perc)) +
  geom_boxplot(fill = "lightcoral") +
  labs(title = "Boxplot of F_perc", y = "F_perc (%)")

p1
p2
```

------------------------------------------------------------------------

## 5. Bivariate Analysis

### F_score vs IMDB Rating

```{r}
ggplot(df, aes(x = Imdb_rating, y = F_score)) +
  geom_point(color = "darkblue", size = 2) +
  geom_smooth(method = "lm", se = FALSE, color = "red") +
  labs(title = "F_score vs IMDb Rating", x = "IMDb Rating", y = "F_score")
```

------------------------------------------------------------------------

## 6. Categorical Flags

```{r}
df %>%
  group_by(Dating_flag) %>%
  summarise(mean_f = mean(F_score, na.rm = TRUE),
            mean_imdb = mean(Imdb_rating, na.rm = TRUE))

df %>%
  group_by(Coaching_flag) %>%
  summarise(mean_f = mean(F_score, na.rm = TRUE),
            mean_imdb = mean(Imdb_rating, na.rm = TRUE))
```

------------------------------------------------------------------------

## 7. Time Series View (by Episode Order)

```{r}
ggplot(df, aes(x = Episode_order, y = F_score)) +
  geom_line(color = "purple", size = 1) +
  geom_point() +
  labs(title = "F_score Over Episode Order", x = "Episode Order", y = "F_score")
```

------------------------------------------------------------------------

## 8. Summary

This EDA highlights: - The distribution and variation in Roy Kent’s `F_score` and `F_perc`. - How these scores vary with `IMDb rating`, and across flags like `Dating_flag` and `Coaching_flag`. - The pattern of Roy’s intensity across the season chronology.

# PART 2

### Hypotheses

1.  **H₁:** Roy's f-bomb count (`F_count_RK`) affects the episode's IMDb rating (`Imdb_rating`).
2.  **H₂:** T he higher the percentage of profanities (`F_perc`), the greater the chance that a dating thread is present (`Dating_flag = "Yes"`).

### Packages, seeding and data preparation

```{r}
df <- df |>
  mutate(
    Season        = factor(Season),
    Dating_flag   = factor(Dating_flag,  levels = c("No", "Yes")),
    Coaching_flag = factor(Coaching_flag, levels = c("No", "Yes"))
  )
```

### Train/test split + CV folders

```{r}
data_split <- initial_split(df, prop = 0.80, strata = Season)
train_data <- training(data_split)
test_data  <- testing(data_split)

cv_folds <- vfold_cv(train_data, v = 5, strata = Season)
```

### Model 1 – linear regression (IMDb)

```{r}
lin_rec <- 
  recipe(Imdb_rating ~ F_count_RK + F_perc + Season +
                          Dating_flag + Coaching_flag,
         data = train_data) |>
  step_dummy(all_nominal_predictors(), one_hot = TRUE) |>
  step_normalize(all_numeric_predictors())
```

```{r}
lin_spec <- linear_reg() |> set_engine("lm")

lin_wf <- workflow() |>
  add_recipe(lin_rec) |>
  add_model(lin_spec)
```

```{r}
lin_res <- fit_resamples(
  lin_wf,
  resamples = cv_folds,
  metrics   = metric_set(rmse, rsq)
)
collect_metrics(lin_res)
```

```{r}
lin_fit  <- fit(lin_wf, data = train_data)

test_pred_lm <- predict(lin_fit, test_data) |>
  bind_cols(test_data |> select(Imdb_rating))

metrics(test_pred_lm, truth = Imdb_rating, estimate = .pred)
tidy(lin_fit)             
```

**Why this model?**

The target (IMDb_rating) is numeric, so ordinary least squares gives the simplest, fully interpretable baseline inside tidymodels.

**Which variables and why?**

-   F_count_RK & F_perc - profanity intensity (hypothesized driver).

-   Season dummies - capture structural shifts across seasons.

-   Dating_flag and Coaching_flag - control for story lines that EDA suggested may alter viewer reception.

**How to interpret the coefficients?**

-   Intercept = 8.32 - baseline rating for a Season 3 episode with no dating/coaching subplot and zero f-bombs.

-   F_count_RK = +0.35 pts per extra f-bomb (p = 0.13, not significant).

-   F_perc ≈ 0 and all dummy variables have small, non-significant effects (p ≫ 0.05). Thus no predictor passes the 0.05 threshold.

**How well does the model fit?**

Cross-validation: RMSE = 0.764, R² = 0.35 (explains \~35 % of in-fold variance). Test set: RMSE = 0.757, R² = 0.002 - almost no explanatory power on unseen data. MAE = 0.631. The model predicts within ≈ 0.76 points but barely beats the mean baseline, indicating weak generalisation.

### Model 2 - Logistic regression (Dating_flag)

```{r}
log_rec <- 
  recipe(Dating_flag ~ F_perc + F_count_RK + Season + Coaching_flag,
         data = train_data) |>
  step_dummy(all_nominal_predictors(), one_hot = TRUE) |>
  step_normalize(all_numeric_predictors())
```

```{r}
log_spec <- logistic_reg() |> set_engine("glm")

log_wf <- workflow() |>
  add_recipe(log_rec) |>
  add_model(log_spec)
```

```{r}
log_res <- fit_resamples(
  log_wf,
  resamples = cv_folds,
  metrics   = metric_set(roc_auc, accuracy)
)
collect_metrics(log_res)
```

```{r}
log_fit <- fit(log_wf, data = train_data)

test_pred_glm <- predict(log_fit, test_data, type = "prob") |>
  bind_cols(test_data |> select(Dating_flag))

roc_auc(test_pred_glm, truth = Dating_flag, .pred_Yes)
tidy(log_fit, exponentiate = TRUE)
```

**Why GLM-logit?**

Because Dating_flag is binary. A logistic GLM models log-odds, outputs class probabilities, and keeps the workflow fully interpretable in tidymodels.

**Odds ratio for F_perc**

Exponentiating the coefficient gives an OR ≈ 1 (the numeric 6.7 × 10⁻⁹³ with p ≈ 0.998 is just a rounding artefact). Changing profanity percentage has no statistically reliable effect on the odds that a dating subplot appears.

**Metrics – what they mean**

-   Accuracy (0.75 in CV) – share of episodes correctly classified; easy to read but can hide class-imbalance issues.

-   AUC (0.93 CV, 0.13 test) – quality of probability ranking; near-1.0 in CV says the model separates classes well in training folds, but the crash to 0.13 on the test set signals severe over-fitting or label inversion, i.e. poor generalisation despite the decent accuracy figure.

### Model comparison

```{r}
mod_compare <- tibble(
  Model        = c("Linear (Imdb_rating)", "Logistic (Dating_flag)"),
  Main_metric  = c(
    paste0("RMSE = ", round(metric_set(rmse)(test_pred_lm, truth = Imdb_rating, estimate = .pred)$.estimate, 3)),
    paste0("AUC = ",  round(roc_auc(test_pred_glm, truth = Dating_flag, .pred_Yes)$.estimate, 3))
  ),
  Interpretability = c("β – pkt IMDb", "OR – odds ratio")
)
mod_compare
```

### Summary

Import & cleaning – loaded richmondway.csv, fixed types, removed NAs, created tidy factors to ensure consistent downstream modelling.

EDA – key findings – profanity is weakly, positively correlated with IMDb score; season shifts and subplot flags show no clear patterns.

Hypotheses → methods – H₁ (Imdb ∼ profanity) → linear regression; H₂ (Dating_flag ∼ profanity) → logistic regression.

Model 1: linear – CV RMSE ≈ 0.76, R²_train ≈ 0.35, R²_test ≈ 0.00; no significant coefficients → profanity/flags don’t explain ratings.

Model 2: logit – CV accuracy ≈ 0.75, AUC ≈ 0.93 but test AUC ≈ 0.13; odds-ratios ≈ 1, no significant effects → model overfits, no real signal.

Comparison – both models are transparent but fail to generalise; language intensity and simple plot flags are insufficient predictors.



