
---
title: "Exploratory Data Analysis (EDA): RichmondWay Dataset Case Study"
author: "Jakub Świerczyński"
date: "`r Sys.Date()`"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE)
```

# Introduction

This notebook conducts an **Exploratory Data Analysis (EDA)** on the `richmondway.csv` dataset, which contains episode-level statistics related to the character Roy Kent in a TV series.

---

## 1. Data Loading and Inspection

```{r}
library(tidyverse)
library(skimr)
library(visdat)

df <- read_csv("richmondway.csv")

# Basic inspection
dim(df)
glimpse(df)
skim(df)
```

---

## 2. Missing Values Overview

```{r fig.width=8, fig.height=5}
vis_miss(df, cluster = TRUE) +
  labs(title = "Missing Value Patterns in RichmondWay Dataset")
```

---

## 3. Data Cleaning and Preparation

```{r}
# Ensure appropriate types
df <- df %>%
  mutate(
    Season = as.integer(Season),
    Episode = as.integer(Episode),
    F_score = as.numeric(F_score),
    F_perc = as.numeric(F_perc),
    year = 2020 + Season # Example: Assigning artificial years if needed
  )

# Check duplicates
sum(duplicated(df))
```

---

## 4. Univariate Analysis

### Distribution of F_score

```{r fig.width=10}
p1 <- ggplot(df, aes(x = F_score)) +
  geom_histogram(bins = 15, fill = "skyblue", color = "black") +
  labs(title = "Histogram of F_score", x = "F_score", y = "Frequency")

p2 <- ggplot(df, aes(y = F_score)) +
  geom_boxplot(fill = "skyblue") +
  labs(title = "Boxplot of F_score", y = "F_score")

p1
p2
```

### Distribution of F_perc

```{r fig.width=10}
p1 <- ggplot(df, aes(x = F_perc)) +
  geom_histogram(bins = 15, fill = "lightcoral", color = "black") +
  labs(title = "Histogram of F_perc", x = "F_perc (%)", y = "Frequency")

p2 <- ggplot(df, aes(y = F_perc)) +
  geom_boxplot(fill = "lightcoral") +
  labs(title = "Boxplot of F_perc", y = "F_perc (%)")

p1
p2
```

---

## 5. Bivariate Analysis

### F_score vs IMDB Rating

```{r}
ggplot(df, aes(x = Imdb_rating, y = F_score)) +
  geom_point(color = "darkblue", size = 2) +
  geom_smooth(method = "lm", se = FALSE, color = "red") +
  labs(title = "F_score vs IMDb Rating", x = "IMDb Rating", y = "F_score")
```

---

## 6. Categorical Flags

```{r}
df %>%
  group_by(Dating_flag) %>%
  summarise(mean_f = mean(F_score, na.rm = TRUE),
            mean_imdb = mean(Imdb_rating, na.rm = TRUE))

df %>%
  group_by(Coaching_flag) %>%
  summarise(mean_f = mean(F_score, na.rm = TRUE),
            mean_imdb = mean(Imdb_rating, na.rm = TRUE))
```

---

## 7. Time Series View (by Episode Order)

```{r}
ggplot(df, aes(x = Episode_order, y = F_score)) +
  geom_line(color = "purple", size = 1) +
  geom_point() +
  labs(title = "F_score Over Episode Order", x = "Episode Order", y = "F_score")
```

---

## 8. Summary

This EDA highlights:
- The distribution and variation in Roy Kent’s `F_score` and `F_perc`.
- How these scores vary with `IMDb rating`, and across flags like `Dating_flag` and `Coaching_flag`.
- The pattern of Roy’s intensity across the season chronology.

