---
title: "11E - Tidying Data with tidyr"
author: "Piotr Kosowski"
date: "2025-03-27"
editor_options: 
  markdown: 
    wrap: sentence
---

# Tidying Data with tidyr

Data **tidying** is a fundamental step in the data analysis process.
Ensuring that your data is in a **consistent, tidy format** makes it easier to manipulate, visualize, and model.
The **`tidyr`** package, part of the **tidyverse**, provides a suite of functions designed to help you **transform messy data** into tidy data.

This notebook covers:

1.  Installing and loading required libraries\
2.  Understanding what constitutes *tidy data*\
3.  Core tidyr functions – `pivot_longer()`, `pivot_wider()`, `separate()`, `unite()`\
4.  Advanced tidyr functions – `fill()`, `replace_na()`, `complete()`\
5.  Practical examples with `nycflights13` data\
6.  Exercises for hands-on practice

------------------------------------------------------------------------

## Setting Up the Environment

Before diving into data tidying, ensure the necessary libraries are installed and loaded.
We will use:

-   **tidyverse** (includes `tidyr`, `dplyr`, `ggplot2`, etc.)\
-   **nycflights13** (for real-world flight data examples)

```{r}
# Define the required libraries
libs <- c("tidyverse", "nycflights13")

# Check if libraries are installed; install missing ones
installed_libs <- libs %in% rownames(installed.packages())
if (any(installed_libs == FALSE)) {
    install.packages(libs[!installed_libs])
}

# Load the libraries
library(tidyverse)
library(nycflights13)
```

**Explanation:**\
- `libs` is a vector of package names.\
- We check if each is installed; if not, we install them.\
- Then we load them with `library()`.

------------------------------------------------------------------------

## Understanding Tidy Data

Tidy data follows a specific structure that makes data analysis more straightforward.
According to **Hadley Wickham**, there are three interrelated rules that make a dataset tidy:

1.  **Each variable forms a column.**\
2.  **Each observation forms a row.**\
3.  **Each type of observational unit forms a table.**

### Why Tidy Data?

**Advantages** of having tidy data:

1.  **Consistency**: A standardized data format simplifies the learning and application of data manipulation tools across the tidyverse (e.g., `dplyr`, `ggplot2`, `tidyr`).\
2.  **Efficiency**: Tidy data leverages R’s vectorized operations, resulting in more concise, readable, and faster-running code.\
3.  **Integration**: Many R packages assume data is “tidy.” You can seamlessly pass tidy data between modeling, plotting, and wrangling functions.

------------------------------------------------------------------------

## Core tidyr Functions

The `tidyr` package offers several functions that help **reshape** and **tidy** data.
The **core** functions include:

1.  **`pivot_longer()`**: Converts wide data to long (“gathering” columns).\
2.  **`pivot_wider()`**: Converts long data to wide (“spreading” columns).\
3.  **`separate()`**: Splits a single column into multiple columns.\
4.  **`unite()`**: Combines multiple columns into one.

We will demonstrate each function with examples – some from the built-in `tidyr` examples like `billboard`, others can be your own data or from `nycflights13`.

------------------------------------------------------------------------

### `pivot_longer()`

`pivot_longer()` transforms data from **wide** format to **long** format, turning multiple column headers into **key-value** pairs.

#### Example: Tidying the `billboard` Dataset

The `billboard` dataset records the billboard rank of songs in the year 2000.\
Each row represents a song, and each `wk#` column represents that song’s rank in a specific week.

```{r}
# Display the first few rows of the billboard dataset
billboard
```

**Observations**:\
- Variables include `artist`, `track`, `date.entered`, plus columns `wk1` through `wk76` for each weekly rank.\
- We need to transform all `wk##` columns into two columns: “week” (the name) and “rank” (the value).

**Pivoting Longer**:

```{r}
# Tidy the billboard dataset by pivoting longer
billboard_tidy <- billboard %>% 
  pivot_longer(
    cols = starts_with("wk"), 
    names_to = "week", 
    values_to = "rank",
    values_drop_na = TRUE
  ) %>% 
  mutate(
    week = parse_number(week)
  )

# Display the tidied dataset
billboard_tidy
```

**Explanation**:

-   `cols = starts_with("wk")`: selects all columns whose names start with `"wk"`.\
-   `names_to = "week"`: The original column names (`wk1`, `wk2`, etc.) are stored in a new column `week`.\
-   `values_to = "rank"`: The values from those columns go into a new column `rank`.\
-   `values_drop_na = TRUE`: Any rows with `NA` in `rank` are removed.\
-   `mutate(week = parse_number(week))`: Converts `"wk1", "wk2", ..."` into numeric `1, 2, ...`.

------------------------------------------------------------------------

#### Many variables in column names

A more challenging scenario arises when column names contain **multiple variables**.
For example, the `who2` dataset:

```{r}
who2
```

**`who2`** records tuberculosis data.
There are columns `country` and `year`, plus \~56 columns like `sp_m_014`, `ep_m_4554`, `rel_m_3544`, etc.
Each column name is made of three pieces separated by `_`:\
1.
The method of diagnosis (`sp`, `rel`, `ep`).\
2.
The gender (`m` or `f`).\
3.
The age range (`014`, `1524`, `2535`, `3544`, `4554`, or `65`).

Hence we have multiple variables embedded in the column name.
We can fix this with:

```{r}
who2_long <- who2 %>% 
  pivot_longer(
    cols = !(country:year),
    names_to = c("diagnosis", "gender", "age"), 
    names_sep = "_",
    values_to = "count"
  )

who2_long
```

**Explanation**:

-   `cols = !(country:year)`: Pivot all columns except `country` through `year`.\
-   `names_to = c("diagnosis", "gender", "age")`: We want to split the column name into these three pieces.\
-   `names_sep = "_"`: The pieces are separated by an underscore in the column name.\
-   `values_to = "count"`: The numeric cell values become `count`.

Alternatively, if the pattern is more complex, we can use `names_pattern` with a regex approach.

------------------------------------------------------------------------

#### Data and variable names in the column headers

Another advanced scenario is when column names mix variable names **and** variable values.
For instance, the `household` dataset might have columns like `name_1`, `dob_1`, `name_2`, `dob_2`, which hold the **child** number (1 or 2) plus the type of data (`name`, `dob`).
We can parse that with `names_to = c(".value", "child")`:

```{r}
household
```

```{r}
household_long <- household %>%
  pivot_longer(
    cols = !family,
    names_to = c(".value", "child"),
    names_sep = "_",
    values_drop_na = TRUE
  ) %>%
  mutate(child = parse_number(child))

household_long
```

The `.value` sentinel indicates that the first part of the column name is an actual new column name, not just data.

------------------------------------------------------------------------

### `pivot_wider()`

`pivot_wider()` is the inverse of `pivot_longer()`.
It spreads key-value pairs across multiple columns, going from a **long** to a **wide** format.

**Example**: The `cms_patient_experience` dataset:

```{r}
cms_patient_experience
```

Might have columns like `org_id`, `measure_cd`, and `prf_rate`.
We can pivot so each measure code becomes a separate column:

```{r}
cms_wide <- cms_patient_experience %>% 
  pivot_wider(
    id_cols = starts_with("org"),
    names_from = measure_cd,
    values_from = prf_rate
  )

cms_wide
```

**Explanation**:

-   `id_cols`: columns that uniquely identify each row (often IDs).\
-   `names_from = measure_cd`: each unique measure code becomes a column name.\
-   `values_from = prf_rate`: the numeric values to place in those columns.

------------------------------------------------------------------------

### `separate()`

`separate()` splits a single column into multiple columns based on a separator.
By default, it splits on **non-alphanumeric** characters, but you can specify `sep`.

**Example**:

```{r}
df <- tibble(x = c(NA, "x.y", "x.z", "y.z"))

df
```

```{r}
df_separated <- df %>%
  separate(x, into = c("A", "B"))

df_separated
```

**Explanation**:

-   If a row is `"x.y"`, after separating, `A = "x"`, `B = "y"`.\
-   `NA` remains `NA` for both A and B.\
-   `separate(..., sep = "_")` would explicitly separate on underscores if needed.

**Handling partial splits**:

```{r}
df_separated_skip <- df %>%
  separate(x, into = c(NA, "B"))  # discard first piece

df_separated_skip
```

------------------------------------------------------------------------

### `unite()`

`unite()` is the opposite of `separate()`.
It combines multiple columns into a single column by concatenating their string values.

**Example**:

```{r}
df <- expand_grid(x = c("a", NA), y = c("b", NA))

df
```

```{r}

df_united <- df %>%
  unite("z", x:y, remove = FALSE, na.rm = TRUE)

df_united
```

**Explanation**:

-   We combine `x` and `y` into a new column `z`.\
-   `na.rm = TRUE` discards `NA` from the concatenation.\
-   `remove = FALSE` keeps the original columns.

------------------------------------------------------------------------

## Advanced tidyr Functions

For more complex situations or missing data scenarios, tidyr provides:

-   **`fill()`**: Fill down (or up) missing values.\
-   **`replace_na()`**: Replace `NA` with a specified value or set of values.\
-   **`complete()`**: Ensure all combinations of specified variables exist, filling missing ones with `NA` or another value.

### `fill()`

Carries forward (or backward) non-missing values:

```{r}
df_fill <- tibble(
  group = c("A", "A", "A", "B", "B", "B"),
  value = c(1, NA, 3, NA, 5, NA)
)

df_fill
```

```{r}

df_filled <- df_fill %>%
  group_by(group) %>%
  fill(value, .direction = "down") %>% 
  ungroup()

df_filled
```

```{r}
df_filled <- df_fill %>%
  group_by(group) %>%
  fill(value, .direction = "downup") %>% 
  ungroup()

df_filled
```

------------------------------------------------------------------------

### `replace_na()`

Replaces `NA` with custom values:

```{r}
df_replace_na <- tibble(
  name = c("Alice", "Bob", "Charlie", "David"),
  score = c(85, NA, 78, NA)
)

df_replace_na
```

```{r}
df_replaced <- df_replace_na %>%
  replace_na(list(score = 0))

df_replaced
```

------------------------------------------------------------------------

### `complete()`

Ensures all combinations of specified variables appear:

```{r}
df_complete <- tibble(
  group = c("A", "A", "B"),
  category = c("X", "Y", "X"),
  value = c(10, 20, 30)
)

df_complete
```

```{r}

df_fully_complete <- df_complete %>%
  complete(group, category)

df_fully_complete
```

------------------------------------------------------------------------

## Exercises

### Exercise 1: Create a Tidy Dataframe

**Question:** Create a dataframe `df` with the following columns: - `id`: Integers from 1 to 5 - `name`: Characters "A", "B", "C", "D", "E" - `score1`, `score2`, `score3`: Random integers between 50 and 100 (use `sample()` function)

```{r}

df <- tibble(
  id = c(1, 2, 3, 4, 5),
  name = c("A", "B", "C", "D", "E"),
  score1 = c(sample(50:100, 5)),
  score2 = c(sample(50:100, 5)),
  score3 = c(sample(50:100, 5))
)
df
```

### Exercise 2: Pivot Longer

**Question:** Transform the dataframe `df` from wide to long format using `pivot_longer()`, gathering the `score1`, `score2`, and `score3` columns into two new columns: `test` and `score`.

```{r}

df_long <- df %>% 
  pivot_longer(
    cols = starts_with("score"),
    names_to = "test",
    values_to = "score"
  )
df_long
```

### Exercise 3: Pivot Wider

**Question:** Using the `df_long` dataframe from the previous exercise, pivot it back to wide format using `pivot_wider()`, spreading the `test` column into separate `score1`, `score2`, and `score3` columns.

```{r}


```

### Exercise 4: Separate Columns

**Question:** Create a dataframe `df_dates` with a column `datetime` containing date and time in the format `"YYYY-MM-DD HH:MM:SS"` (just type proper values inside the `c()` function).
Split this column into two separate columns: `date` and `time`.

```{r}


```

**Separating the `datetime` Column:**

```{r}


```

### Exercise 5: Unite Columns

**Question:** Using the `df_separated` dataframe from the previous exercise, combine the `date` and `time` columns back into a single `datetime` column using `unite()`.

```{r}



```

### Exercise 6: Handling Missing Values with fill()

**Question:** Create a dataframe `df_missing` with `id` and `score` columns with some missing values in the `score` column (use `tibble()` function) .
Use `fill()` to propagate the last non-missing value forward.

```{r}


```

**Filling Missing Values:**

```{r}


```

### Exercise 7: Replacing NAs with Specific Values

**Question:** Using the `df_missing` dataframe, replace all `NA` values in the `score` column with the average score.

```{r}



```

### Exercise 8: Complete Missing Combinations

**Question:** Create a dataframe `df_complete` with `id` and `category` columns, but missing some combinations.
Use `complete()` to ensure all combinations are present, filling missing `value` with `0`.

```{r}



```

**Completing the Dataframe:**

```{r}



```

### Exercise 9: Advanced Pivoting with Multiple Variables

**Question:** Create a dataframe `df_multi` where column names contain multiple variables separated by underscores (e.g., `type_gender_age`).
Use `separate()` to split these into individual variables.

```{r}



```

**Separating Multiple Variables:**

```{r}



```

### Exercise 10: Combining tidyr and dplyr for Data Analysis

**Question:** Using the `flights` dataset from `nycflights13`, perform the following steps: 1.
Filter flights that departed in January.
2.
Select relevant columns: `year`, `month`, `day`, `dep_time`, `arr_time`, `carrier`, `tailnum`.
3.
Separate the `dep_time` and `arr_time` into `dep_hour` and `dep_minute`, `arr_hour` and `arr_minute`.
4.
Unite `dep_hour` and `dep_minute` into `departure_time`, and `arr_hour` and `arr_minute` into `arrival_time`.
5.
Replace any `NA` in `departure_time` or `arrival_time` with `"00:00"`.

```{r}



```

------------------------------------------------------------------------

## Summary

In this session, you explored **tidying data** with the **`tidyr`** package:

-   **Core functions** like `pivot_longer()`, `pivot_wider()`, `separate()`, and `unite()` to reshape data.\
-   **Advanced functions** (`fill()`, `replace_na()`, `complete()`) for dealing with missing or partial data.\
-   How these transformations align with the **tidy data principles** of having one variable per column and one observation per row.

**Key Points**:

-   Tidy data is essential for a smooth analytical workflow in R.\
-   `pivot_longer()` and `pivot_wider()` help you reshape data between wide and long formats.\
-   `separate()` and `unite()` help split or combine column values.\
-   Functions like `fill()`, `replace_na()`, and `complete()` handle missing or incomplete data.\
-   Integrating tidyr with **dplyr** allows for robust data cleaning pipelines.

**Next Steps**:

-   **Practice**: Apply tidyr transformations to real-world messy datasets.\
-   **Combine**: Use tidyr with `dplyr` for data manipulation, `ggplot2` for visualization.\
-   **Explore**: Look at advanced arguments like `names_pattern` (regular expressions) in `pivot_longer()` and more.\
-   **Real-world projects**: Tidy data is critical for reproducible, scalable data analyses.

------------------------------------------------------------------------

# References and Additional Resources

-   [**tidyr Documentation**](https://tidyr.tidyverse.org/)\
-   [**R for Data Science** (Hadley Wickham & Garrett Grolemund)](https://r4ds.had.co.nz/)\
-   [**Advanced R** (Hadley Wickham)](https://adv-r.hadley.nz/)\
-   [**Tidyverse Cheatsheets**](https://www.rstudio.com/resources/cheatsheets/)\
-   [**Swirl**: Interactive R tutorials, including data tidying topics](https://swirlstats.com/students.html)
