---
title: "10E - dplyr"
author: "Piotr Kosowski"
date: "2024-10-07"
editor_options: 
  markdown: 
    wrap: sentence
---

# dplyr: Data Manipulation in R

The `dplyr` package, part of the tidyverse collection, is a powerful tool for data manipulation in R.
It provides a set of functions (verbs) that are intuitive and efficient for transforming and summarizing data.
This notebook will delve into the fundamentals of `dplyr`, explore both pipe operators, and introduce additional essential functions to enhance your data manipulation skills.

------------------------------------------------------------------------

## Setting Up the Environment

Before diving into data manipulation, ensure that the necessary libraries are installed and loaded.
We will use the `tidyverse` package, which includes `dplyr` among other essential packages, and the `nycflights13` dataset for practical examples.

```{r}
# Define the required libraries
libs <- c("tidyverse", "nycflights13")

# Check if libraries are installed; install missing ones
installed_libs <- libs %in% rownames(installed.packages())
if (any(installed_libs == FALSE)) {
    install.packages(libs[!installed_libs])
} 

# Load the libraries
library(tidyverse)
library(nycflights13)
```

**Explanation:** - **`libs`**: A vector containing the names of required libraries.
- **Installation Check**: Verifies if each library is installed; installs any that are missing.
- **`library()`**: Loads the specified libraries into the R session.

------------------------------------------------------------------------

## Understanding Pipe Operators

Pipe operators allow for more readable and concise code by passing the output of one function directly into the next function.
In R, there are two primary pipe operators:

1.  **Magrittr Pipe (`%>%`)**:
    -   **Origin**: Comes from the `magrittr` package, which is loaded with `tidyverse`.
    -   **Usage**: Widely used in the tidyverse for chaining multiple operations.
    -   **Example**: `r      df %>%         filter(condition) %>%         arrange(column)`
2.  **Native Pipe (`|>`)**:
    -   **Origin**: Introduced in R version 4.1.0 as part of base R.
    -   **Usage**: Similar to the magrittr pipe but with some syntactical differences.
    -   **Example**: `r      df |>         filter(condition) |>         arrange(column)`

### Key Differences Between `%>%` and `|>`

-   **Function Naming**:
    -   `%>%`: Allows for more flexible function placement, including anonymous functions.
    -   `|>`: More rigid, primarily designed for simple function chaining.
-   **Backward Compatibility**:
    -   `%>%`: Part of the magrittr package and used extensively in existing tidyverse code.
    -   `|>`: Native to R and will be more standard in future R code.
-   **Anonymous Functions**:
    -   `%>%`: Supports the dot (`.`) placeholder for more complex operations.
    -   `|>`: Requires explicit anonymous function definitions for similar complexity.

### Examples

#### Using `%>%` Pipe

```{r}
# Example using magrittr pipe to filter and arrange data
filtered_arranged <- flights %>% 
  filter(dest == "IAH") %>% 
  arrange(desc(arr_delay))

# Display the result
filtered_arranged
```

#### Using `|>` Pipe

```{r}
# Example using native pipe to filter and arrange data
filtered_arranged_native <- flights |> 
  filter(dest == "IAH") |> 
  arrange(desc(arr_delay))

# Display the result
filtered_arranged_native
```

**Explanation:** - Both examples achieve the same outcome: filtering flights destined for "IAH" and arranging them in descending order of arrival delay.
- The difference lies in the pipe operator used (`%>%` vs. `|>`).

------------------------------------------------------------------------

## dplyr Basics

`dplyr` provides a set of functions (verbs) that facilitate data manipulation.
The primary verbs include:

1.  **filter()**: Selects rows based on conditions.
2.  **arrange()**: Orders rows based on column values.
3.  **distinct()**: Identifies unique rows.
4.  **mutate()**: Adds or modifies columns.
5.  **select()**: Chooses specific columns.
6.  **rename()**: Renames columns.
7.  **relocate()**: Changes the position of columns.
8.  **pull()**: Extracts a single column as a vector.
9.  **group_by()**: Groups data for summarization.
10. **summarize()**: Creates summary statistics.
11. **join functions**: Combines multiple data frames.

### Common Characteristics of dplyr Verbs

-   **First Argument**: Always a data frame.
-   **Subsequent Arguments**: Describe actions based on column names.
-   **Output**: Always a new data frame.

These characteristics make it straightforward to chain multiple operations using pipe operators.

------------------------------------------------------------------------

### filter()

`filter()` allows you to retain rows that meet certain conditions.

**Syntax:**

``` r
filter(.data, condition1, condition2, ...)
```

**Examples:**

#### Example 1: Flights with Arrival Delay \> 120 Minutes

```{r}
# Filter flights that arrived more than 120 minutes late
delayed_flights <- flights %>% 
  filter(arr_delay > 120)

# Display the result
delayed_flights

```

#### Example 2: Flights on January 1st

```{r}

# Filter flights that departed on January 1st
january_first_flights <- flights %>% 
  filter(month == 1 & day == 1)

# Display the result
january_first_flights

```

#### Example 3: Flights Departed in January or February

```{r}

# Filter flights that departed in January or February
winter_flights <- flights %>% 
  filter(month %in% c(1, 2))

# Display the result
winter_flights

```

**Explanation:** - **`arr_delay > 120`**: Selects flights with arrival delays exceeding two hours.
- **`month == 1 & day == 1`**: Selects flights on January 1st.
- **`month %in% c(1, 2)`**: Selects flights in January or February using the `%in%` operator for brevity.

------------------------------------------------------------------------

### arrange()

`arrange()` orders the rows of a data frame based on the values of specified columns.

**Syntax:**

``` r
arrange(.data, column1, column2, ..., .by_group = FALSE)
```

**Examples:**

#### Example 1: Arrange Flights by Departure Time

```{r}

# Arrange flights by departure time in ascending order
flights_sorted_dep_time <- flights %>% 
  arrange(year, month, day, dep_time)

# Display the result
flights_sorted_dep_time

```

#### Example 2: Arrange Flights by Departure Delay Descending

```{r}

# Arrange flights by descending departure delay
flights_sorted_dep_delay <- flights %>% 
  arrange(desc(dep_delay))

# Display the result
flights_sorted_dep_delay

```

#### Example 3: Arrange Flights by Arrival Delay and Distance

```{r}

# Arrange flights by arrival delay descending and then by distance
flights_sorted_complex <- flights %>% 
  arrange(desc(arr_delay), distance) 

# Display the result
flights_sorted_complex

```

**Explanation:** - **`arrange(year, month, day, dep_time)`**: Sorts flights chronologically based on departure time.
- **`desc(dep_delay)`**: Sorts flights by departure delay in descending order.
- **`arrange(desc(arr_delay), distance)`**: First sorts by arrival delay descending, then by distance ascending to break ties.

------------------------------------------------------------------------

### distinct()

`distinct()` identifies and retains unique rows based on specified columns.

**Syntax:**

``` r
distinct(.data, column1, column2, ..., .keep_all = FALSE)
```

**Examples:**

#### Example 1: Remove Duplicate Rows

```{r}

# Remove duplicate rows from the flights dataset
unique_flights <- flights %>% 
  distinct()

# Display the result
unique_flights

```

#### Example 2: Unique Origin-Destination Pairs

```{r}

# Find unique origin and destination pairs
unique_routes <- flights %>% 
  distinct(origin)

# Display the result
unique_routes

```

#### Example 3: Count of Unique Destinations

```{r}

# Count the number of unique destinations
unique_dest_count <- flights %>% 
  distinct(dest) %>% 
  count()

# Display the result
unique_dest_count

```

**Explanation:** - **`distinct()`**: By default, removes duplicate rows based on all columns.
- **`distinct(origin, dest)`**: Retains unique combinations of `origin` and `dest`.
- **`count()`**: Counts the number of unique entries after `distinct()`.

------------------------------------------------------------------------

## Manipulating Columns

Manipulating columns is essential for shaping data frames to fit analysis needs.
`dplyr` offers several verbs for this purpose: `mutate()`, `select()`, `rename()`, `relocate()`, and `pull()`.

### mutate()

`mutate()` adds new columns or modifies existing ones based on computations or transformations.

**Syntax:**

``` r
mutate(.data, new_column = expression, ...)
```

**Examples:**

#### Example 1: Calculate Gain and Speed

```{r}

# Add 'gain' and 'speed' columns to the flights dataset
flights_mutated <- flights %>% 
  mutate(
    gain = dep_delay - arr_delay,
    speed = distance / air_time * 60
  )

# Display the result
flights_mutated

```

**Explanation:** - **`gain`**: Represents how much time a flight made up (positive) or lost (negative) during the flight.
- **`speed`**: Calculates the speed in miles per hour based on distance and air time.

#### Example 2: Adding Columns Before a Specific Position

```{r}

# Add 'gain' and 'speed' columns before the first column
flights_mutated_before <- flights %>% 
  mutate(
    gain = dep_delay - arr_delay,
    speed = distance / air_time * 60,
    .before = 1
  )

# Display the result
flights_mutated_before

```

**Explanation:** - **`.before = 1`**: Places the new columns at the beginning of the data frame.
- **`.after = column_name`**: Similarly, you can specify a column after which to place new columns.

#### Example 3: Keeping Only Used Columns

```{r}

# Add 'gain', 'hours', and 'gain_per_hour' while keeping only used columns
flights_mutated_keep <- flights %>% 
  mutate(
    gain = dep_delay - arr_delay,
    hours = air_time / 60,
    gain_per_hour = gain / hours,
    .keep = "used"
  )

# Display the result
flights_mutated_keep

```

**Explanation:** - **`.keep = "used"`**: Retains only the columns used in the `mutate()` operations alongside the new columns.

------------------------------------------------------------------------

### select()

`select()` allows you to choose specific columns from a data frame, making it easier to focus on relevant variables.

**Syntax:**

``` r
select(.data, column1, column2, ..., helpers)
```

**Helpers:** - **`starts_with("prefix")`**: Selects columns that start with the specified prefix.
- **`ends_with("suffix")`**: Selects columns that end with the specified suffix.
- **`contains("text")`**: Selects columns that contain the specified text.
- **`matches("regex")`**: Selects columns that match the regular expression.
- **`everything()`**: Selects all remaining columns.

**Examples:**

#### Example 1: Select Specific Columns

```{r}

# Select only 'year', 'month', and 'day' columns
selected_columns <- flights %>% 
  select(year, month, day)

# Display the result
selected_columns

```

#### Example 2: Select Columns Using Range

```{r}

# Select all columns from 'year' to 'day' inclusive
selected_range <- flights %>% 
  select(year:day)

# Display the result
selected_range

```

#### Example 3: Exclude Specific Columns

```{r}

# Select all columns except from 'year' to 'day' inclusive
excluded_columns <- flights %>% 
  select(!year:day)

# Display the result
excluded_columns

```

#### Example 4: Select Columns Based on Data Type

```{r}

# Select all character columns
character_columns <- flights %>% 
  select(where(is.character))

# Display the result
character_columns

```

#### Example 5: Select Columns Using Helper Functions

```{r}

# Select columns that start with 'dep'
dep_columns <- flights %>% 
  select(starts_with("dep"))

# Display the result
dep_columns

```

**Explanation:** - **`select(year, month, day)`**: Retains only the specified columns.
- **`select(year:day)`**: Selects a range of columns from `year` to `day`.
- **`select(!year:day)`**: Excludes the specified range of columns.
- **`select(where(is.character))`**: Selects columns with the character data type.
- **`starts_with("dep")`**: Selects all columns that begin with "dep", such as `dep_time`, `dep_delay`.

------------------------------------------------------------------------

### rename()

`rename()` changes the names of existing columns without altering the data.

**Syntax:**

``` r
rename(.data, new_name = old_name, ...)
```

**Examples:**

#### Example 1: Rename a Single Column

```{r}

# Rename 'tailnum' to 'tail_num'
renamed_flights <- flights %>% 
  rename(tail_num = tailnum)

# Display the result
renamed_flights

```

#### Example 2: Rename Multiple Columns

```{r}

# Rename 'dep_time' to 'departure_time' and 'arr_time' to 'arrival_time'
renamed_flights_multiple <- flights %>% 
  rename(
    departure_time = dep_time,
    arrival_time = arr_time
  )

# Display the result
renamed_flights_multiple

```

#### Example 3: Rename Columns Using Helper Functions

```{r}

# Rename all columns that start with 'dep' to start with 'departure_'
renamed_dep <- flights %>% 
  rename_with(~ str_replace(., "^dep", "departure"))

# Display the result
renamed_dep

```

**Explanation:** - **`rename(tail_num = tailnum)`**: Changes the column name `tailnum` to `tail_num`.
- **`rename_with()`**: Utilizes a helper function to rename multiple columns based on a pattern.

------------------------------------------------------------------------

### relocate()

`relocate()` changes the position of columns within a data frame, allowing you to organize data more logically.

**Syntax:**

``` r
relocate(.data, column1, column2, ..., .before = column, .after = column)
```

**Examples:**

#### Example 1: Move Columns to the Beginning

```{r}

# Relocate 'time_hour' and 'air_time' to the beginning
relocated_flights <- flights %>% 
  relocate(time_hour, air_time)

# Display the result
relocated_flights

```

#### Example 2: Move Columns After a Specific Column

```{r}

# Relocate 'year' to be after 'time_hour'
relocated_flights_after <- flights %>% 
  relocate(year, .after = time_hour)

# Display the result
relocated_flights_after

```

#### Example 3: Move Columns Using Helper Functions

```{r}

# Relocate columns that start with 'arr' before 'dep_time'
relocated_arr_before_dep <- flights %>% 
  relocate(starts_with("arr"), .before = dep_time)

# Display the result
relocated_arr_before_dep

```

**Explanation:** - **`relocate(time_hour, air_time)`**: Moves the specified columns to the front of the data frame.
- **`.after = time_hour`**: Places the relocated columns immediately after the `time_hour` column.
- **`starts_with("arr")`**: Selects columns that start with "arr" for relocation.

------------------------------------------------------------------------

### pull()

`pull()` extracts a single column from a data frame as a vector, facilitating further analysis or operations.

**Syntax:**

``` r
pull(.data, column)
```

**Examples:**

#### Example 1: Extract 'dep_delay' as a Vector

```{r}

# Extract the 'dep_delay' column as a vector
departure_delays <- flights %>% 
  pull(dep_delay)

# Display the first few elements
head(departure_delays)

```

#### Example 2: Extract 'carrier' Column

```{r}

# Extract the 'carrier' column as a vector
carriers <- flights %>% 
  pull(carrier)

# Display the first few elements
head(carriers)

```

**Explanation:** - **`pull(dep_delay)`**: Extracts the `dep_delay` column, allowing you to perform vectorized operations on it.

------------------------------------------------------------------------

## Grouping and Summarizing Data

Grouping data is essential for performing aggregate operations, such as calculating summaries for different subsets of data.
`dplyr` provides the `group_by()` and `summarize()` functions to facilitate these tasks.

### group_by()

`group_by()` divides the data into groups based on the values of specified columns.
Subsequent operations, such as `summarize()`, will then be performed within these groups.

**Syntax:**

``` r
group_by(.data, column1, column2, ...)
```

**Examples:**

#### Example 1: Group Flights by Month

```{r}

# Group flights by month
flights_grouped_month <- flights %>% 
  group_by(month)

# Display the grouped data
flights_grouped_month

```

#### Example 2: Group Flights by Carrier and Destination

```{r}

# Group flights by carrier and destination
flights_grouped_carrier_dest <- flights %>% 
  group_by(carrier, dest)

# Display the grouped data
flights_grouped_carrier_dest

```

**Explanation:** - **`group_by(month)`**: Groups the flights data by each month.
- **`group_by(carrier, dest)`**: Creates groups based on unique combinations of carrier and destination.

### summarize()

`summarize()` creates summary statistics for each group defined by `group_by()`.

**Syntax:**

``` r
summarize(.data, summary1 = expression, summary2 = expression, ...)
```

**Examples:**

#### Example 1: Average Departure Delay by Month

```{r}

# Calculate average departure delay for each month
avg_dep_delay_month <- flights %>% 
  group_by(month) %>% 
  summarize(
    avg_dep_delay = mean(dep_delay, na.rm = TRUE)
  )

# Display the result
avg_dep_delay_month

```

#### Example 2: Count of Flights per Carrier

```{r}

# Count the number of flights per carrier
flight_count_carrier <- flights %>% 
  group_by(carrier) %>% 
  summarize(
    total_flights = n()
  )

# Display the result
flight_count_carrier

```

#### Example 3: Multiple Summaries for Each Destination

```{r}

# Calculate average arrival delay and total flights per destination
delay_summary_dest <- flights %>% 
  group_by(dest) %>% 
  summarize(
    avg_arr_delay = mean(arr_delay, na.rm = TRUE),
    total_flights = n()
  )

# Display the result
delay_summary_dest

```

**Explanation:** - **`mean(dep_delay, na.rm = TRUE)`**: Calculates the average departure delay, excluding missing values.
- **`n()`**: Counts the number of rows (flights) in each group.

### slice\_\* Functions

The `slice_*` family of functions allows you to select specific rows within each group, such as the first or last row, or rows with minimum/maximum values.

**Functions:** - **`slice_head(n = 1)`**: Selects the first `n` rows of each group.
- **`slice_tail(n = 1)`**: Selects the last `n` rows of each group.
- **`slice_min(order_by, n = 1)`**: Selects the rows with the smallest values in `order_by`.
- **`slice_max(order_by, n = 1)`**: Selects the rows with the largest values in `order_by`.
- **`slice_sample(n = 1)`**: Selects random rows from each group.

**Examples:**

#### Example 1: Most Delayed Flight to Each Destination

```{r}

# Find the most delayed flight to each destination
most_delayed_dest <- flights %>% 
  group_by(dest) %>% 
  slice_max(arr_delay, n = 1)

# Display the result
most_delayed_dest

```

#### Example 2: First Flight of Each Day

```{r}

# Find the first flight of each day
first_flight_daily <- flights %>% 
  group_by(year, month, day) %>% 
  slice_head(n = 1)

# Display the result
first_flight_daily

```

#### Example 3: Random Sample of Flights per Carrier

```{r}

# Select a random sample of 2 flights per carrier
random_flights_carrier <- flights %>% 
  group_by(carrier) %>% 
  slice_sample(n = 2)

# Display the result
random_flights_carrier

```

**Explanation:** - **`slice_max(arr_delay, n = 1)`**: Retrieves the flight with the maximum arrival delay for each destination.
- **`slice_head(n = 1)`**: Gets the first flight of each day based on the current ordering.
- **`slice_sample(n = 2)`**: Randomly selects two flights per carrier.

### count()

`count()` provides a quick way to count the number of occurrences of combinations of variables.

**Syntax:**

``` r
count(.data, column1, column2, ..., sort = FALSE, name = "n")
```

**Examples:**

#### Example 1: Count of Flights per Origin-Destination Pair

```{r}

# Count the number of flights for each origin-destination pair
flight_counts_od <- flights %>% 
  count(origin, dest, sort = TRUE)

# Display the result
flight_counts_od

```

#### Example 2: Count of Flights per Carrier

```{r}

# Count the number of flights per carrier
flight_counts_carrier <- flights %>% 
  count(carrier, sort = TRUE)

# Display the result
flight_counts_carrier

```

**Explanation:** - **`count(origin, dest, sort = TRUE)`**: Counts flights for each unique combination of `origin` and `dest`, sorted in descending order of count.
- **`count(carrier, sort = TRUE)`**: Counts flights per carrier, sorted from most to least.

------------------------------------------------------------------------

## Additional dplyr Functions

While the primary verbs cover most data manipulation needs, `dplyr` offers additional functions that enhance data processing capabilities.

### Join Functions

Joins combine two data frames based on a common key.
The primary join functions include:

1.  **left_join()**: Keeps all rows from the left data frame and adds matching rows from the right.
2.  **right_join()**: Keeps all rows from the right data frame and adds matching rows from the left.
3.  **inner_join()**: Keeps only rows that have matching keys in both data frames.
4.  **full_join()**: Keeps all rows from both data frames, filling in `NA` where there are no matches.

**Example: Joining Flights with Airlines Data**

```{r}

# View the airlines dataset
airlines

```

```{r}

# Perform a left join to add airline names to flights data
flights_with_airlines <- flights %>% 
  left_join(airlines, by = "carrier")

# Display the result
flights_with_airlines

```

**Explanation:** - **`left_join(airlines, by = "carrier")`**: Merges the `airlines` data frame into `flights` based on the `carrier` column, retaining all flights and adding airline names where available.

### bind_rows() and bind_cols()

These functions combine data frames by rows or columns.

**Examples:**

#### bind_rows()

```{r}

# Create two small data frames
df1 <- data.frame(id = 1:3, value = c("A", "B", "C"))
df2 <- data.frame(id = 4:6, value = c("D", "E", "F"))

# Bind the rows together
combined_rows <- bind_rows(df1, df2)

# Display the result
combined_rows

```

#### bind_cols()

```{r}

# Create two small data frames with the same number of rows
df3 <- data.frame(id = 1:3, value = c("A", "B", "C"))
df4 <- data.frame(score = c(90, 85, 88))

# Bind the columns together
combined_cols <- bind_cols(df3, df4)

# Display the result
combined_cols

```

**Explanation:** - **`bind_rows()`**: Stacks `df1` and `df2` vertically, combining their rows.
- **`bind_cols()`**: Places `df3` and `df4` side by side, combining their columns.

### rename_with()

`rename_with()` allows you to rename multiple columns using a function, enabling bulk renaming based on patterns or transformations.

**Syntax:**

``` r
rename_with(.data, .fn, .cols = everything())
```

**Examples:**

#### Convert Column Names to Uppercase

```{r}

# Convert all column names to uppercase
flights_upper <- flights %>% 
  rename_with(toupper)

# Display the result
flights_upper

```

#### Replace Underscores with Dashes in Column Names

```{r}

# Replace underscores with dashes in column names
flights_renamed <- flights %>% 
  rename_with(~ str_replace_all(., "_", "-"))

# Display the result
flights_renamed

```

**Explanation:** - **`rename_with(toupper)`**: Transforms all column names to uppercase.
- **`rename_with(~ str_replace_all(., "_", "-"))`**: Replaces all underscores with dashes in column names using a lambda function.

### transmute()

`transmute()` creates new columns while dropping existing ones, unlike `mutate()` which retains all existing columns.

**Syntax:**

``` r
transmute(.data, new_column = expression, ...)
```

**Examples:**

#### Calculate Gain and Speed, Drop Other Columns

```{r}

# Create a data frame with only 'gain' and 'speed' columns
flights_transmuted <- flights %>% 
  transmute(
    gain = dep_delay - arr_delay,
    speed = distance / air_time * 60
  )

# Display the result
flights_transmuted

```

**Explanation:** - **`transmute()`**: Adds new columns `gain` and `speed` while removing all other columns from the data frame.

------------------------------------------------------------------------

## Exercises

### Using Built-in `mtcars` Dataset

The `mtcars` dataset is a built-in R dataset that comprises fuel consumption and 10 aspects of automobile design and performance for 32 automobiles (1973–74 models).

#### Exercise 1: Filter Cars with mpg \> 20 and 6 Cylinders

**Question:** Return rows of cars that have an `mpg` value greater than 20 and 6 cylinders.

```{r}
mtcars %>% 
  filter(mpg > 20 & cyl == 6)

```

#### Exercise 2: Reorder the Data Frame by `cyl` and Descending `wt`

**Question:** Reorder the `mtcars` data frame first by the number of cylinders (`cyl`), then by descending weight (`wt`).

```{r}
mtcars <- mtcars %>% 
  arrange(cyl, desc(wt))

```

#### Exercise 3: Select the `mpg` and `hp` Columns

**Question:** Select only the `mpg` (miles per gallon) and `hp` (horsepower) columns from the `mtcars` dataset.

```{r}
mtcars %>% 
  select(mpg, hp)

```

#### Exercise 4: Select Distinct Values of the `gear` Column

**Question:** Select the distinct values of the `gear` column from the `mtcars` dataset.

```{r}

mtcars %>% 
  select(gear) %>% 
  distinct()

```

#### Exercise 5: Create a New Column `Performance` as `hp` Divided by `wt`

**Question:** Create a new column called `Performance` which is calculated by horsepower (`hp`) divided by weight (`wt`).

```{r}

mtcars <- mtcars %>% 
  mutate(
    Performance = hp / wt
  )


```

#### Exercise 6: Calculate the Average Horsepower for Each Number of Cylinders

**Question:** Create a summary table that shows the average horsepower (`hp`) for each number of cylinders (`cyl`).

```{r}

mtcars %>% 
  group_by(cyl) %>% 
  summarise(
    avg_hp = mean(hp)
  )


```

#### Exercise 7: Identify Cars with the Highest and Lowest MPG

**Question:** Identify the car(s) with the highest and lowest miles per gallon (`mpg`).

```{r}

mtcars %>%
  filter(mpg == max(mpg) | mpg == min(mpg))


```

```{r}



```

#### Exercise 8: Add a Column Indicating Transmission Type

**Question:** Add a new column `transmission` that labels cars as "Automatic" if `am` is 0 and "Manual" if `am` is 1.

```{r}


```

#### Exercise 9: Select Cars with High Performance

**Question:** Select cars with a `Performance` value greater than 30, where `Performance` is calculated as `hp / wt`.

```{r}


```

#### Exercise 10: Rename Columns for Clarity

**Question:** Rename the `hp` column to `horsepower` and `wt` to `weight`.

```{r}


```

------------------------------------------------------------------------

### Using `flights` Dataset from `nycflights13`

The `flights` dataset contains information about all flights that departed from New York City in 2013.

#### Exercise 1: Find All Flights That Had an Arrival Delay of Two or More Hours

**Question:** Find all flights that had an arrival delay of two or more hours.

```{r}



```

#### Exercise 2: Find Flights That Flew to Houston (`IAH` or `HOU`)

**Question:** Find all flights that flew to Houston, either `IAH` or `HOU`.

```{r}



```

#### Exercise 3: Find Flights Operated by United, American, or Delta

**Question:** Find all flights operated by United, American, or Delta airlines.

```{r}



```

#### Exercise 4: Find Flights Departed in Summer (July, August, September)

**Question:** Find all flights that departed in the summer months: July, August, and September.

```{r}


```

#### Exercise 5: Find Flights That Arrived More Than Two Hours Late but Didn’t Leave Late

**Question:** Find flights that arrived more than two hours late but didn’t leave late (departure delay \<= 0).

```{r}



```

#### Exercise 6: Find Flights Delayed by at Least an Hour but Made Up Over 30 Minutes in Flight

**Question:** Find flights that were delayed by at least an hour upon departure but made up over 30 minutes during the flight (arrival delay \<= departure delay - 30).

```{r}



```

#### Exercise 7: Sort `flights` to Find the Flights with Longest Departure Delays and Earliest Departure Times

**Question:** Sort `flights` to find the flights with the longest departure delays and those that left earliest in the morning.

```{r}


```

```{r}


```

#### Exercise 8: Sort `flights` to Find the Fastest Flights

**Question:** Sort `flights` to find the fastest flights by calculating speed (`distance / air_time * 60`).

```{r}


```

#### Exercise 9: Check if There Was a Flight on Every Day of 2013

**Question:** Verify if there was at least one flight every day in 2013.

```{r}


```

#### Exercise 10: Find the Most and Least Distant Flights

**Question:** Identify which flights traveled the farthest and the least distance.

```{r}



```

```{r}


```

#### Exercise 11: Determine Which Carrier Has the Worst Delays

**Question:** Identify the carrier with the worst average arrival delays.

```{r}



```

------------------------------------------------------------------------

#### Exercise 12: Find the Most Delayed Flight to Each Destination

**Question:** Find the most delayed flight to each destination.

```{r}



```

#### Exercise 13: Analyze How Delays Vary Over the Course of the Day

**Question:** Analyze how departure and arrival delays vary over different times of the day.

```{r}



```

#### Exercise 14: Calculate Total Number of Flights per Day

**Question:** Calculate the total number of flights for each day in 2013.

```{r}


```

#### Exercise 15: Find the Most Frequent Destinations

**Question:** Identify the top 10 most frequent destination airports.

```{r}



```

#### Exercise 16: Calculate Average Speed per Carrier

**Question:** Calculate the average speed (miles per hour) for each carrier.

```{r}


```

#### Exercise 17: Identify Days with the Highest Number of Canceled Flights

**Question:** Identify the days with the highest number of canceled flights (`is.na(dep_time)`).

```{r}



```

#### Exercise 18: Analyze Departure Delays by Carrier

**Question:** Compare the average departure delays among different carriers.

```{r}



```

------------------------------------------------------------------------

## Summary

In this session, we explored the `dplyr` package, a cornerstone of the tidyverse for data manipulation in R.
We covered the essential verbs that allow for filtering, arranging, selecting, mutating, and summarizing data.
Additionally, we delved into more advanced functions such as joins, binding rows/columns, and renaming with helper functions.

Understanding pipe operators is crucial for writing readable and efficient code.
We compared the magrittr pipe (`%>%`) with the native R pipe (`|>`), highlighting their differences and use cases.

Through practical examples and exercises using the `mtcars` and `flights` datasets, you have gained hands-on experience in applying `dplyr` functions to real-world data scenarios.
These skills are foundational for any data analysis workflow in R, enabling you to transform and summarize data effectively.

**Next Steps:** - **Practice**: Implement more complex data manipulation tasks using `dplyr` to reinforce your understanding.
- **Explore**: Dive deeper into advanced functions and explore the `tidyr` package for data tidying.
- **Visualization**: Combine `dplyr` with visualization packages like `ggplot2` to create insightful plots.
- **Optimization**: Learn about optimizing data manipulation for larger datasets and leveraging parallel processing where applicable.

------------------------------------------------------------------------

# References and Additional Resources

-   **R Documentation: dplyr**: Comprehensive documentation for all `dplyr` functions.
    -   [dplyr on CRAN](https://CRAN.R-project.org/package=dplyr)
-   **R for Data Science by Hadley Wickham & Garrett Grolemund**: An excellent resource for learning data science in R, covering `dplyr` in depth.
    -   [R for Data Science](https://r4ds.had.co.nz/)
-   **Advanced R by Hadley Wickham**: For a deeper understanding of functions and programming concepts in R.
    -   [Advanced R](https://adv-r.hadley.nz/)
-   **dplyr Cheat Sheet**: Quick reference for `dplyr` functions and syntax.
    -   [dplyr Cheat Sheet](https://rstudio.com/resources/cheatsheets/)
